Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2016-05-13 cikm2016/nlp.tex
minor edits to nlp section.

The second algorithm is a natural language approach to parsing date and time formats. While regular expressions can be used to match strings to known formats, we chose a grammar approach as a generalization of regular expressions. Such an approach provides for greater expressibility and less rigidity, without the expectation of the input string to exactly match against one or more of the known formats. We describe a particular form of grammar called a context-free grammar approach as well as extensions to the rules of the grammar that are inherent to date and time properties such as grammar variants, constraints, and assigning probabilistic weighting through a supervised learning approach.

\subsection{Context-Free Grammar}
ICU date formats are well defined both structurally and semantically, and can be defined by a context-free grammar (CFG). This has several advantages. First, it allows for modular syntax definitions, which simplifies grammar development and enables reuse. Second, it grants total freedom in structuring a grammar to best fit its intended use. The power of such a grammar is its expressiveness, and is not constrained by pattern rigidities that are prevalent with regular expressions and text processing~\cite{Grune:1990}.

@@

A classical CFG performs poorly on inflected expressions resulting from misapplication of morphological inflection and syntactic rules. Large numbers of non-terminals are necessary for representation of all variations of such features, and the parser output would consist of many parse trees. We extend the date-time grammar by adding morpho-syntactic variants. These variants allows the parser to correct for syntactic errors (\textit{e.g.} usage of whitespace characters and punctuation marks), capitalization errors and abbreviations (\textit{e.g.} \texttt{`Mon'} for \texttt{Monday}). We use several various external corpora for such corrections~\cite{nltkcorpora}.

@@

\item \textbf{LeapYear Restriction for \texttt{February}}:
While the above restriction applies to all months barring \texttt{February}, we also apply a constraint to the number of days for \texttt{February}, based on whether the year is leap year or not. For this, we define a new symbol in the grammar called \texttt{LeapYear}. If an expression containing the month \texttt{`February'} or any such variant (\textit{e.g.} \texttt{`Feb'}, \texttt{`2'}) with the day \texttt{`29'} and a year, would need to resolve the \texttt{Year} symbol to be a \texttt{LeapYear}, defined as $\texttt{Year} \Mod 4 == 0$. However, this is just an approximation. The Gregorian calendar also requires that a year evenly divisible by $100$ (\textit{e.g.}, 1800) is a leap year only if it is also evenly divisible by $400$.

@@

The grammar parser's success is often limited by data ambiguity, and incomplete expressions. For example an input string \texttt{`3/7/2005'} could either be interpreted as as the 7th day of March or the 3rd day of July depending on the date format used. While a single entry cannot resolve the parser's ambiguity, a column of data provides more context for determining the dominant pattern prevalent in the dataset.

>>>>

The second algorithm is a natural language approach to parsing date and time formats. While regular expressions can be used to match strings to known formats, we chose a grammar approach as a generalization of regular expressions. Such an approach provides for greater expressibility and less rigidity, without the expectation of the input string to \textit{exactly} match against one or more of the known formats. We describe a particular form of grammar called a context-free grammar approach as well as extensions to the rules of the grammar that are inherent to date and time properties such as grammar variants, constraints, and assigning probabilistic weighting through a supervised learning approach.

\subsection{Context-Free Grammar}
ICU date formats are well defined both structurally and semantically, and can be defined by a context-free grammar (CFG). This has several advantages. First, it allows for modular syntax definitions, which simplifies grammar development and enables reuse. Second, it grants total freedom in structuring a grammar to best fit its intended use. The power of such a grammar is its expressiveness, and is not constrained by pattern rigidities that are prevalent in regular expressions and text processing~\cite{Grune:1990}.

@@

A classical CFG performs poorly on inflected expressions resulting from misapplication of morphological inflection and syntactic rules. Large numbers of non-terminals are necessary for representation of all variations of such features, and the parser output would consist of many parse trees. We extend the date-time grammar by adding morpho-syntactic variants. These variants allows the parser to correct for syntactic errors (\textit{e.g.}, usage of whitespace characters and punctuation marks), capitalization errors and abbreviations (\textit{e.g.}, \texttt{`Mon'} for \texttt{Monday}). We use several various external corpora for such corrections~\cite{nltkcorpora}.

@@

\item \textbf{LeapYear Restriction for \texttt{February}}:
While the above restriction applies to all months barring \texttt{February}, we also apply a constraint to the number of days for \texttt{February}, based on whether the year is leap year or not. For this, we define a new symbol in the grammar called \texttt{LeapYear}. If an expression containing the month \texttt{February} or any such variant (\textit{e.g.}, \texttt{`Feb'}, \texttt{`2'}) with the day \texttt{`29'} and a year, would need to resolve the \texttt{Year} symbol to be a \texttt{LeapYear}, defined as $\texttt{Year} \Mod 4 == 0$. However, this is just an approximation. The Gregorian calendar also requires that a year evenly divisible by $100$ (\textit{e.g.}, 1800) is a leap year only if it is also evenly divisible by $400$.

@@

The parser's success is often limited by data ambiguity, and incomplete expressions. For example an input string \texttt{`3/7/2005'} could either be interpreted as as the 7th day of March or the 3rd day of July depending on the date format used. While a single entry cannot resolve the parser's ambiguity, a column of data provides more context for determining the dominant pattern prevalent in the dataset.
